Synthesis of Linear Functional Programs

### 2021-05-06 article/article.tex
First introduction iteration

% General motivation?
Program synthesis is an automated or semi-automated process of
 deriving a program (generating code) from a high-level specification.
 Specifications appear in many forms (i.e.
 polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
 examples~\cite{DBLP:conf/popl/FrankleOWZ16},
 graded linear types~\cite{DBLP:conf/lopstr/HughesO20}, ...),
 and in this work the synthesis specification will be a linear type, ?possibly with refinements.?
 The chosen specification form dictates the synthesis process, and different ones might
 be approached in completely different ways.
 In spite of the specification, most program synthetizer's have to deal with two main
 inherent complexities: searching over a vast space of valid programs, and interpreting user intent.
 Synthesis has many motivations, it can be seen as a means to improve a programmer's productivity and/or
 program correctness (i.e. through suggestion and/or autocompletion), or as a tool to automate certain parts
 of the programming process (i.e. in the same way AI might generate some boilerplate text to auxiliate a journalist, rather
 than to replace), just to name a few examples.

% Type based synthesis
Type-based synthesis takes a given type and produces a program with that same type.

This work explores the process of synthetising linear functional
programs from types based in linear logic, leveraging the Curry-Howard correspondence.
Said correspondence states that propositions in a logic have a direct mapping to types,
 and well typed programs also have a one to one correspendence to proofs of those propositions.
As such, having a type be a proposition in linear logic, we can relate a proof of that
proposition directly to a linear functional program
 — finding a proof is finding a program with that type, and so we
 formulate the synthesis as proof search in linear logic.

>>>>

% General motivation?
Program synthesis is an automated or semi-automated process of
 deriving a program (generating code) from a high-level specification.
 Specifications appear in many forms (e.g.
 polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
 examples~\cite{DBLP:conf/popl/FrankleOWZ16},
 graded linear types~\cite{DBLP:conf/lopstr/HughesO20}, ...).
 In this work the synthesis specification will be a linear type, ?possibly with refinements.?
 The chosen specification form dictates the synthesis process, and different ones might
 be approached in completely different ways.
 In spite of the specification, most program synthetizer's have to deal with two main
 inherent complexities: searching over a vast space of valid programs, and interpreting user intent.
 Synthesis has many motivations, it can be seen as a means to improve a programmer's productivity and/or
 program correctness (i.e. through suggestion and/or autocompletion), or as a tool to automate certain parts
 of the programming process (e.g. in the same way AI might generate some boilerplate text to auxiliate a journalist, rather
 than to replace), just to name a few examples.

% Type based synthesis
Type-based synthesis takes a type and produces a program with that given type. 
Types can be further explored to give the programmer more expressiveness (e.g. by adding constraints??), and are called rich types.
-- Gostava de ver com o prof aqui. [Estive a procurar e fiquei um pouco confuso. polimorfismo é rich types?
dependent types are rich types? aqueles constraints são os dependent types pelo que percebi... mas n me
parece correto. arithmetic types? temos de rever. Por agora vou por apenas os tópicos que o professor escreveu]
Languages with richer type systems allow for more precise
  types. This precision can statically eliminate various kinds of
  logical errors by making certain invalid program states ill-typed
  (e.g. a ``null aware'' type system, will ensure you cannot
  dereference a null-pointer).
However, they can also be a burden -- the whole point of these type
  systems is to ensure that ``less'' things are well-typed, and
  sometimes it's hard to convince the checker.
Regarding the challenges, type-based synthesis leverages rich types as a way of pruning
  the search space, and by using types gives the user as a more ``familiar'' specification.

This work explores the process of synthetising linear functional
 programs from types based in linear logic (linear types), leveraging the Curry-Howard correspondence.
Said correspondence states that propositions in a logic have a direct mapping to types,
 and well typed programs also have a one to one correspendence to proofs of those propositions.
As such, having a type be a proposition in linear logic, we can relate a proof of that
 proposition directly to a linear functional program
 — finding a proof is finding a program with that type, so we
 formulate the synthesis as proof search in linear logic, which makes to us available
 a lot of related technology from the proof search literature.
Linear types differ from usual types for they constrain resource usage in programs by
 \emph{statically?} limiting the number of times certain resources can be used during their lifetime.
 They can be applied to resource-aware programming such as concurrent programming (e.g. session types??), and to memory-management (e.g. Rust)
 [Quero reescrever a frase acima, não gosto de como falamos de session types e rust em dois exemplos paralelos visto que não são coisas bem comparáveis. Revemos na reunião.]

 % Goals
In the end, we intend to be able to do full and partial synthesis of well-typed programs --
 Full synthesis being the production of a function (or set of) satisfying the specification;
 Partial synthesis being the ``completion'' of a partial program (i.e. a function with a \emph{hole} in it) --
 Starting from a small core linear -- i.e. \emph{resource-aware} functional language, and building up to recursive types/functions.
 Eventually, possibly, enhancing the synthesis process with type refinements, and making it interactive -- i.e. the synthetiser offers the user
 ``choices'' mid-process.
[Devia estar a escrever isto eu ahah?]The evaluation will be done through expressiveness
benchmarks (i.e. can we synthesize ``X''?); time measurements (i.e. how fast can we synthesize ``X''?'); and so on.
